---
title: "CSDE 502 Winter 2022, Assignment 9"
author: "mbwaters"
contact: "mbwaters@uw.edu"
output: 
    bookdown::html_document2:
        number_sections: true
        self_contained: true
        code_folding: hide
        toc: true
        toc_float:
            collapsed: true
            smooth_scroll: false
    pdf_document:
        number_sections: true
        toc: true
        fig_cap: yes
        keep_tex: yes
urlcolor: blue 
---

<!-- This is a CSS style sheet for the answers -->
<style>
h1 {
  font-size: 22px;}
  
h2 {
  font-size: 18px;}
  
h3 {
  font-size: 18px;}

p {
  font-size: 16px;}

.answer {
  font-size: 18px;
  background-color: lightblue;
  border: 2px solid red;
  border-radius: 12px;
  padding: 5px;  
  } 
  
.indent {
  background-color: lightblue;
  padding-left: 50px;
  
  } 
}
</style>

```{r}
library(captioner)

figure_nums <- captioner(prefix = "Figure")
table_nums <- captioner(prefix = "Table")

# path to this file name
if (!interactive()) {
    fnamepath <- knitr::current_input(dir = TRUE)
} else {
    fnamepath <- ""
}
```

***Explanation***:
This assignment is intended to give you more practice delving into the Add Health data set and in manipulating additional variables. 

***Instructions***: 

1. Make sure your Rmd file has no local file system dependencies (i.e., anyone should be able to recreate the output HTML using only the Rmd source file).
1. Make a copy of this Rmd file and add answers below each question. The code that generated the answers should be included, as well as the complete source code for the document.\    
Put your answers within the **$\lt$div class="answer"$\gt$...$\lt$/div$\gt$** tags so they will show up within the blue boxes with red borders.
1. Change the YAML header above to identify yourself and include contact information.
1. For any tables or figures, include captions and cross-references and any other document automation methods as necessary.
1. Make sure your output HTML file looks appealing to the reader.
1. Upload the final Rmd to your github repository.
1. Download [`assn_id.txt`](http://staff.washington.edu/phurvitz/csde502_winter_2021/assignments/assn_id.txt) and include the URL to your Rmd file on github.com.
1. Create a zip file from your copy of `assn_id.txt` and upload the zip file to the Canvas site for Assignment 1. ***The zip file should contain only the text file. Do not include any additional files in the zip file--everything should be able to run from the file you uploaded to github.com. Please use zip format and not 7z or any other compression/archive format.***


#
**Using the full household roster (you'll need to go back the full raw data source, [21600-0001-Data.dta](http://staff.washington.edu/phurvitz/csde502_winter_2021/data/21600-0001-Data.dta.zip)), create the following variables for each respondent. Document any decisions that you make regarding missing values, definitions, etc. in your narrative as well as in the R code.  Include a frequency tabulation and a histogram of each result.**

<div class="answer">
Answer
</div>

##
**Total number in household**

<div class="answer">
```{r}
#install and load tidyverse and haven
install.packages("tidyverse", repos = "http://cran.us.r-project.org")
install.packages("haven", repos = "http://cran.us.r-project.org")
library(tidyverse)
library(haven)
mytempdir <- tempdir()

# download and unzip the larger Add Health data set
myUrl <- "http://staff.washington.edu/phurvitz/csde502_winter_2021/data/21600-0001-Data.dta.zip"
zipfile <- file.path(tempdir(), basename(myUrl))
dtafile <- tools::file_path_sans_ext(zipfile)

# check if the dta file exists
if (!file.exists(dtafile)) {
    # if the dta file doesn't exist, check for the zip file
    # check if the zip file exists, download if necessary
    if (!file.exists(zipfile)) {
        curl::curl_download(url = myUrl, destfile = zipfile)
    }
    # unzip the downloaded zip file
    if (file.exists(zipfile)) {
        unzip(zipfile = zipfile, exdir = tempdir())
    }
}

# if the data set has not been read, read it in
if (!exists("ahcomplete")) {
    ahcomplete <- haven::read_dta(dtafile)
}

# make the data frame by selecting columns that begin with H1HR2
#choose HIHR2 because any HIHR variable could be chosen but you only need 1 (from HIHR2-HIHR11, household member # is notated by the letter at the end of the variable name)
ah_h1hr2 <- ahcomplete %>%
    # select columns
    select(AID, starts_with("H1HR2")
)
#recode values to 1 or 0, to later sum
#responses of male (1) or female (2) or refused (6) were counted as 1 (presence of household member), all other responses coded as 0
ah_h1hr2$AID <- as.numeric(ah_h1hr2$AID)
recode2 <- function(x, na.rm = FALSE){if_else(x <= 6, 1, 0, missing = NULL)}
ah_h1hr2_2 <- ah_h1hr2 %>% mutate_at(vars(contains('H1HR2')), recode2)
#add up by row
ah_h1hr2_2 <- ah_h1hr2_2 %>% rowwise() %>% mutate(total = sum(c_across(where(is.numeric))))
#subtract AID from total
ah_h1hr2_2 <- ah_h1hr2_2 %>% mutate(total = total - AID)
#freq tab
household_freq <- ah_h1hr2_2 %>% group_by(total) %>% summarise(freq = n())
household_freq %>%
  knitr::kable(caption = "Frequency tabulation of household members") %>%
  kableExtra::kable_classic()
```
```{r}
#histogram
ggplot(ah_h1hr2_2, aes(total)) + geom_histogram() + theme_bw() + xlab("Total household members")
```

</div>

##
**Number of sisters**

<div class="answer">
```{r}
ah_h1hr3 <- ahcomplete %>%
    # select columns
    select(AID, starts_with("H1HR3")
)
#recode values to 1 or 0, to later sum
#responses of sister (8) where coded as 1 and all others 0
ah_h1hr3$AID <- as.numeric(ah_h1hr3$AID)
recode3 <- function(x, na.rm = FALSE){if_else(x == 8, 1, 0, missing = NULL)}
ah_h1hr3_3 <- ah_h1hr3 %>% mutate_at(vars(contains('H1HR3')), recode3)
#add up by row
ah_h1hr3_3 <- ah_h1hr3_3 %>% rowwise() %>% mutate(total = sum(c_across(where(is.numeric))))
#subtract AID from total
ah_h1hr3_3 <- ah_h1hr3_3 %>% mutate(total = total - AID)
#freq tab
sisters_freq <- ah_h1hr3_3 %>% group_by(total) %>% summarise(freq = n())
sisters_freq %>%
  knitr::kable(caption = "Frequency tabulation of sisters") %>%
  kableExtra::kable_classic()
```

```{r}
#histogram
ggplot(ah_h1hr3_3, aes(total)) + geom_histogram() + theme_bw() + xlab("Total number of sisters")
```

</div>

##
**Number of brothers**

<div class="answer">
```{r}
#recode values to 1 or 0, to later sum
#responses of brother (5) coded as 1, all else 0
recode4 <- function(x, na.rm = FALSE){if_else(x == 5, 1, 0, missing = NULL)}
ah_h1hr3_4 <- ah_h1hr3 %>% mutate_at(vars(contains('H1HR3')), recode4)
#add up by row
ah_h1hr3_4 <- ah_h1hr3_4 %>% rowwise() %>% mutate(total = sum(c_across(where(is.numeric))))
#subtract AID from total
ah_h1hr3_4 <- ah_h1hr3_4 %>% mutate(total = total - AID)
#freq tab
brothers_freq <- ah_h1hr3_4 %>% group_by(total) %>% summarise(freq = n())
brothers_freq %>%
  knitr::kable(caption = "Frequency tabulation of brothers") %>%
  kableExtra::kable_classic()
```

```{r}
#histogram
ggplot(ah_h1hr3_4, aes(total)) + geom_histogram() + theme_bw() + xlab("Total number of brothers")
```

</div>

##
**Total number of siblings**

<div class="answer">
```{r}
#recode values to 1 or 0, to later sum
#responses of siblings
recode5 <- function(x, na.rm = FALSE){if_else(x == 5 | x == 8, 1, 0, missing = NULL)}
ah_h1hr3_5 <- ah_h1hr3 %>% mutate_at(vars(contains('H1HR3')), recode5)
#add up by row
ah_h1hr3_5 <- ah_h1hr3_5 %>% rowwise() %>% mutate(total = sum(c_across(where(is.numeric))))
#subtract AID from total
ah_h1hr3_5 <- ah_h1hr3_5 %>% mutate(total = total - AID)
#freq tab
siblings_freq <- ah_h1hr3_5 %>% group_by(total) %>% summarise(freq = n())
siblings_freq %>%
  knitr::kable(caption = "Frequency tabulation of siblings") %>%
  kableExtra::kable_classic()
```

```{r}
#histogram
ggplot(ah_h1hr3_5, aes(total)) + geom_histogram() + theme_bw() + xlab("Total number of siblings")
```

</div>

#
**What proportion of students live with two biological parents? Include the analysis in your R code.**

<div class="answer">
```{r}
#recode values to 1 or 0, to later sum
#responses of siblings
recode6 <- function(x, na.rm = FALSE){if_else(x == 11 | x == 14, 1, 0, missing = NULL)}
ah_h1hr3_6 <- ah_h1hr3 %>% mutate_at(vars(contains('H1HR3')), recode5)
#add up by row
ah_h1hr3_6 <- ah_h1hr3_6 %>% rowwise() %>% mutate(total = sum(c_across(where(is.numeric))))
#subtract AID from total
ah_h1hr3_6 <- ah_h1hr3_6 %>% mutate(total = total - AID)
#value >2 in total counted as true
bio_parents_freq <- ah_h1hr3_6 %>% mutate(bio_parents = case_when(total >= 2 ~ 1, total < 2 ~ 0))
```

The proportion of students who live with two biological parents is `r round(sum(bio_parents_freq$bio_parents)/6504, 3)`.

</div>
#
**Calculate the number of household members that are NOT biological mother, biological father, full brother or full sister. Create a contingency table and histogram for this variable.**

<div class="answer">
```{r}
#recode values to 1 or 0, to later sum, all sister, brother, mother, father reponses coded to 0, all else 1 
recode6 <- function(x, na.rm = FALSE){if_else(x == 5 | x == 8 | x == 11 | x == 14 | x == 96 | x == 97 | x == 98 | x == 99, 0, 1, missing = NULL)}
ah_h1hr3_7 <- ah_h1hr3 %>% mutate_at(vars(contains('H1HR3')), recode6)
#add up by row
ah_h1hr3_7 <- ah_h1hr3_7 %>% rowwise() %>% mutate(total = sum(c_across(where(is.numeric))))
#subtract AID from total
ah_h1hr3_7 <- ah_h1hr3_7 %>% mutate(total = total - AID)
#freq tab
non_bio_freq <- ah_h1hr3_5 %>% group_by(total) %>% summarise(freq = n())
non_bio_freq %>%
  knitr::kable(caption = "Frequency tabulation of non-biological household members") %>%
  kableExtra::kable_classic()
```
```{r}
#histogram
ggplot(ah_h1hr3_7, aes(total)) + geom_histogram() + theme_bw() + xlab("Total number of non-biological household members")
```

</div>

<hr>
Rendered at <tt>`r Sys.time()`</tt>

## Source code
File is at `r fnamepath`.

### R code used in this document
```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```

### Complete Rmd code
```{r comment=''}
cat(readLines(fnamepath), sep = '\n')
```
